# Privacy FAQs {#faq}

On this page you can find Frequently Asked Questions (FAQs) about handling 
personal data in research. Click a question you have to read its answer. 

```{r faqsformattingfunction, message=FALSE, warning=FALSE, echo=FALSE,include=FALSE}
# Install and load the readxl package
#install.packages("readxl")
library(readxl)

formatfaqs <- function(faqsheet){
  qanda <- readxl::read_excel("chapters/faqs.xlsx", 
                              sheet = faqsheet)
  
  formattedtext <- list()
  
  # Format each question in html <details> tags
  for(questionitem in 1:dim(qanda)[1]){
    
    tryCatch( { 
      if(qanda$Include[questionitem] == 1){
        formattedtext[[questionitem]] <- paste0("<details><summary><b>",
                                                qanda[questionitem,1],
                                                "</b></summary><div>",
                                                qanda[questionitem,2],
                                                "</div><br></details>")
      }
    }, error = function(e){
      cat("Error in row : ",questionitem,"\n")
    }
    )
  } #endfor
  # Remove empty list items (happens when Include == 0)
  formattedtext <- formattedtext[!sapply(formattedtext,is.null)]

  # Concatenate the list of items into a single string
  formattedtext_string <- paste(formattedtext, collapse="")
  return(formattedtext_string)
}
```

### General questions {#general}

```{r faqs-general, message=FALSE,warning=FALSE,echo=FALSE}
faqs_general <- formatfaqs("faqs-general")
```

`r faqs_general`

### Procedures and responsibilities {#procedures}

```{r faqs-procedures, message=FALSE,warning=FALSE,echo=FALSE}
faqs_procedures <- formatfaqs("faqs-procedures")
```

`r faqs_procedures`

### Informed consent {#consent}

```{r faqs-ic, message=FALSE,warning=FALSE,echo=FALSE}
faqs_ic <- formatfaqs("faqs-ic")
```

`r faqs_ic`

### Legal questions {#legal}

```{r faqs-legal, message=FALSE,warning=FALSE,echo=FALSE}
faqs_legal <- formatfaqs("faqs-legal")
```

`r faqs_legal`

### Storing personal data {#storage}

```{r faqs-storage, message=FALSE,warning=FALSE,echo=FALSE}
faqs_storage <- formatfaqs("faqs-storage")
```

`r faqs_storage`

### Sharing, publishing and reusing personal data {#sharing}

```{r faqs-share, message=FALSE,warning=FALSE,echo=FALSE}
faqs_share <- formatfaqs("faqs-share")
```

`r faqs_share`

### Practical questions {#practical-questions}

```{r faqs-practical, message=FALSE,warning=FALSE,echo=FALSE}
faqs_practical <- formatfaqs("faqs-practical")
```

`r faqs_practical`

### Students and student data {#students}

```{r faqs-student, message=FALSE,warning=FALSE,echo=FALSE}
faqs_student <- formatfaqs("faqs-student")
```

`r faqs_student`

### Finding support {#finding-support}

```{r faqs-support, message=FALSE,warning=FALSE,echo=FALSE}
faqs_support <- formatfaqs("faqs-support")
```

`r faqs_support`