# Privacy FAQs {#faq}

On this page you can find some often-heard questions about handling personal
data in research. Click a question you have to read its answer.

This is a first FAQ. More will follow!

<details>
  <summary><b>Am I a processor as employee of my university?</b></summary>
  No. As an employee you are still determining your own why (research question) 
  and how (methods) of personal data processing. This makes you a controller, 
  acting as an "agent" of the legal controller (your university). Read more on the
  difference between processors and controllers <a href="#definitions">on the 
  definitions page</a>.
</details>

```{r faqsformattingfunction, message=FALSE, warning=FALSE, echo=FALSE}
# Install and load the readxl package if not already installed
#install.packages("readxl")
library(readxl)

formatfaqs <- function(faqsheet){
  #qanda <- readxl::read_excel("faqs.xlsx", sheet = faqsheet)
  qanda <- readxl::read_excel("chapters/faqs.xlsx", sheet = faqsheet)

  formattedtext <- list()
  
  # Format each question in html <details> tags
  for(questionitem in 1:dim(qanda)[1]){
    formattedtext[[questionitem]] <- paste0("<details><summary><b>",
                                            qanda[questionitem,1],
                                            "</b></summary><div>",
                                            qanda[questionitem,2],
                                            "</div></details>")
  }
  
  # Concatenate the list of items into a single string
  formattedtext_string <- paste(formattedtext, collapse="")
  return(formattedtext_string)
}
```

### General questions

```{r faqs-general, message=FALSE,warning=FALSE,echo=FALSE}
faqs_general <- formatfaqs("faqs-general")
```

`r faqs_general`

### Procedural questions

```{r faqs-procedures, message=FALSE,warning=FALSE,echo=FALSE}
faqs_procedures <- formatfaqs("faqs-procedures")

```

`r faqs_procedures`
