# Privacy FAQs {#faq}
:::keywords
Date of last review: 2023-01-27
:::

On this page you can find Frequently Asked Questions (FAQs) about handling 
personal data in research. Click a question you have to read its answer. 

```{r formatfaqs, message=FALSE, warning=FALSE, echo=FALSE}

# Read in faqs.csv
faqs <- read.csv("files/faqs.csv", stringsAsFactors = FALSE)

# Format FAQs in HTML <details> tags - function is used in chunks below
formatfaqs <- function(faqsection){
  
  # Select the relevant section
  subselection <- faqs[faqs$section_id == faqsection, ]
  
  formattedtext <- list()
  
  # Format each question in the section in html <details> tags
  for(questionitem in 1:nrow(subselection)){
    
    tryCatch( { 
      # Only include if "include" == 1
      if(subselection$include[questionitem] == 1){
        formattedtext[[questionitem]] <- paste0("<details><summary>**",
                                                subselection$question[questionitem],
                                                "**</summary><div>",
                                                subselection$answer[questionitem],
                                                "</div><br></details>")
      }
    }, error = function(e){
      cat("Error in row : ",questionitem,"\n")
    }
    )
  } #end for
  
  # Remove empty list items (happens when Include == 0)
  formattedtext <- formattedtext[!sapply(formattedtext, is.null)]#

  # Concatenate the list of items into a single string
  formattedtext_string <- paste(formattedtext, collapse="")
  return(formattedtext_string)
} # end function
```

### General questions {#general}

```{r faqs-general, message=FALSE,warning=FALSE,echo=FALSE}
faqs_general <- formatfaqs("general")
```

`r faqs_general`

### Procedures and responsibilities {#procedures}

```{r faqs-procedures, message=FALSE,warning=FALSE,echo=FALSE}
faqs_procedures <- formatfaqs("procedures")
```

`r faqs_procedures`

### Informed consent {#consent}

```{r faqs-ic, message=FALSE,warning=FALSE,echo=FALSE}
faqs_ic <- formatfaqs("consent")
```

`r faqs_ic`

### Legal questions {#legal}

```{r faqs-legal, message=FALSE,warning=FALSE,echo=FALSE}
faqs_legal <- formatfaqs("legal")
```

`r faqs_legal`

### Storing personal data {#storage}

```{r faqs-storage, message=FALSE,warning=FALSE,echo=FALSE}
faqs_storage <- formatfaqs("storage")
```

`r faqs_storage`

### Sharing, publishing and reusing personal data {#sharing}

```{r faqs-share, message=FALSE,warning=FALSE,echo=FALSE}
faqs_share <- formatfaqs("sharing")
```

`r faqs_share`

### Practical questions {#practical-questions}

```{r faqs-practical, message=FALSE,warning=FALSE,echo=FALSE}
faqs_practical <- formatfaqs("practical-questions")
```

`r faqs_practical`

### Students and student data {#students}

```{r faqs-student, message=FALSE,warning=FALSE,echo=FALSE}
faqs_student <- formatfaqs("students")
```

`r faqs_student`

### Finding support {#finding-support}

```{r faqs-support, message=FALSE,warning=FALSE,echo=FALSE}
faqs_support <- formatfaqs("finding-support")
```

`r faqs_support`
