# Statistical approaches to de-identification {#statistical-privacy}

In order to protect datasets that contain personal or otherwise sensitive data, 
there is an increasing number of statistical approaches to de-identification, 
which to some extent quantify how identifiable data are after 
[de-identification](#pseudonymisation-anonymisation).  

In this chapter, we discuss the following approaches, as these are the most 
widely used approaches: 

- [K-anonymity](#k-anonymity) 
- [L-diversity](#l-diversity) 
- [T-closeness](#t-closeness) 
- [Differential privacy](#differential-privacy) 

These approaches (or: privacy models) are not yet much used in research 
practice, because they come with some disadvantages and require resources 
and/or expertise to be applied and interpreted correctly. However, they are 
used in many 
[de-identification tools](https://github.com/UtrechtUniversity/privacy-engineering-tools/tree/main/deidentification){target="_blank"} 
and are useful to detect specific sensitivities in (tabular) datasets. For 
those reasons, the techniques are introduced in this chapter. 

## K-anonymity, l-diversity and t-closeness {#k-l-t-anonymity}

:::keywords
On this page: k-anonymous, l-diverse, t-close, privacy model, quantifying 
privacy, key attribute, sensitive attribute, quasi-identifier  
Date of last review: 2023-05-30
:::

K-anonymity, L-diversity and T-closeness are statistical approaches that 
quantify the level of identifiability within a tabular dataset, especially when 
variables within that dataset are combined. They are complementary approaches: 
a dataset can be k-anonymous, L-diverse and T-close, where k, L and T all 
represent a number.  

### Identifiers, quasi-identifiers, and sensitive attributes {#k-anonymity-variables}

Privacy models like k-anonymity, L-diversity and T-closeness distinguish 
between 3 types of variables in a dataset: 

- **Identifiers** (also known as key attributes): direct identifiers such as 
names, student numbers, email addresses, etc. These variables should in 
principle not be collected at all, or removed from the dataset if they are not 
necessary for your research project. 
- **Quasi-identifiers**: indirect identifiers that can lead to identification 
when combined with other quasi-identifiers in the dataset or external 
information. These are often demographic variables like age, sex, place of 
residence, etc., but could also be something entirely different like physical 
characteristics, timestamps, etc. In general, quasi-identifiers are usually 
variables that are likely to be known to someone in the outside world.  
- **Sensitive attributes**: variables of interest which should be protected, and 
which cannot be changed, because they are the main outcome variables. For 
example, it can be Medical condition in a healthcare dataset, or Income in a 
financial dataset. 

It is important to correctly categorise the variables in your dataset as any of 
these variable types if you want to apply k-anonymity, l-diversity and 
t-closeness, because they will determine how the dataset will be de-identified. 

### How it works
#### K-anonymity {#k-anonymity}
K-anonymity ensures that each individual in a dataset cannot be distinguished 
from at least k-1 other individuals with respect to the quasi-identifiers in the 
dataset. This is done through [generalisation](#generalisation), 
[suppression](#suppression) and sometimes [top- and bottom-coding](#top-bottom-coding). 
Applying k-anonymity makes it more difficult for an attacker to re-identify 
specific individuals in the dataset. It protects against 
[singling out and, to some extent, the Mosaic effect](#personal-data-assess). 

```{r k-anonymity-tables, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
# Here is the documentation for kableExtra for html: https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html
# And here it is for LateX/pdf: https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf
library(knitr)
install.packages("kableExtra")
library(kableExtra)

# Sample data frames
df1 <- data.frame(Nr = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), 
                  Age = c(16, 18, 19, 22, 22, 23, 24, 25, 26, 28),
                  Sex = c("Male", "Male", "Male", "Female", "Male", "Male", 
                          "Male", "Female", "Female", "Female"),
                  City = c("Rotterdam", "Rotterdam", "Rotterdam", "Rotterdam", 
                           "Zwolle", "Zwolle", "Zwolle", "Utrecht", 
                           "Rotterdam",  "Utrecht"),
                  Disease = c("Viral infection", "Heart-related", "Cancer", 
                              "Viral infection", "No illness", "Tuberculosis", 
                              "Heart-related", "Cancer", "Heart-related", 
                              "Tuberculosis"))

df2 <- data.frame(Nr = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), 
                  Age = c("=< 20", "=< 20", "=< 20", "20-30", "20-30", "20-30",
                          "20-30",  "20-30",  "20-30",  "20-30"),
                  Sex = c("Male", "Male", "Male", "Female", "Male", "Male", 
                          "Male", "Female", "Female", "Female"),
                  City = c("Rotterdam", "Rotterdam", "Rotterdam", "Rotterdam", 
                           "Zwolle", "Zwolle", "Zwolle", "Utrecht", 
                           "Rotterdam",  "Utrecht"),
                  Disease = c("Viral infection", "Heart-related", "Cancer", 
                              "Viral infection", "No illness", "Tuberculosis", 
                              "Heart-related", "Cancer", "Heart-related", 
                              "Tuberculosis"))

# Convert data frames to tables
table1 <- 
  kbl(df1, caption = "Original dataset", label = NA) %>% 
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 11,
    #fixed_thead = list(enabled = T, background = "#FFCD00")
  ) %>%
  column_spec(2, bold = T) %>%
  row_spec(row = 1:3, background = "#ffedc3") %>%
  row_spec(row = c(4, 9), background = "#b2dfd8") %>%
  row_spec(row = 5:7, background = "#c9e0fc") %>%
  row_spec(row = c(8, 10), background = "#fad6bf") #%>%
#footnote(general = "Original dataset", general_title = "", footnote_as_chunk = T)

table2 <- 
  kbl(df2, caption = "2-anonymous dataset", label = NA) %>% 
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 11,
    #fixed_thead = list(enabled = T, background = "#FFCD00")
  ) %>%
  column_spec(2, bold = T) %>%
  row_spec(row = 1:3, background = "#ffedc3") %>%
  row_spec(row = c(4, 9), background = "#b2dfd8") %>%
  row_spec(row = 5:7, background = "#c9e0fc") %>%
  row_spec(row = c(8, 10), background = "#fad6bf") %>%
  footnote(general = "Colours indicate an 'equivalence class' of quasi-identifers", 
           general_title = "", footnote_as_chunk = T)

if(knitr::is_html_output()){
cat(c(
      '<table><tr valign="top"><td>', 
      table1, 
      '</td><td>', 
      table2, 
      '</td><tr></table>'),
    sep = '')
  
} else if(knitr::is_latex_output()){
  print(table1)
  print(table2)
} else if(knitr::is_epub_output()){
  print(table1)
  print(table2)
}
```

To make a dataset k-anonymous, you must first identify which variables in the 
dataset are identifiers, quasi-identifiers, and sensitive attributes. In the 
example above, Age, Sex and City are quasi-identifiers and Disease is the 
sensitive attribute. Next, you should set a value for k. If we choose a k of 2, 
every row in the example dataset should have the same combination of Age, Sex 
and City as at least 1 other row in the dataset. Finally, you aggregate the 
dataset so that every combination of quasi-identifiers occurs at least k times. 
In the example, this was done by generalising Age into age categories, but 
there may also be other ways to reach 2-anonymity in this dataset. 

There is no single value for k 
[which you should always choose](https://desfontain.es/privacy/k-anonymity.html#how-to-choose-k){target="_blank"}. 
The higher the k, the more difficult it will be to identify someone, but likely 
your dataset will also become less granular and perhaps less informative. The 
value of k will be highly dependent on what you communicated to data subjects 
(e.g., you may have promised a certain k) and the risk of identification that 
you are willing to accept.  

The below [video](https://www.youtube.com/watch?v=Q0DNOIGUzMc){target="_blank"} 
gives an example on how k-anonymity can work in practice: 

<iframe width="100%" height="400" src="https://www.youtube-nocookie.com/embed/Q0DNOIGUzMc?end=189" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>

#### L-diversity {#l-diversity}

L-diversity is an extension to k-anonymity that ensures that there is sufficient 
variation in a *sensitive attribute*. This is important, because if all 
individuals in a (subset of a) dataset have the same value for the sensitive 
attribute, there is still a risk of *inference*. For example, in the below 
2-anonymous dataset, you can infer that any female from Rotterdam between 20 and 
30 who participated had a viral infection ("homogeneity attack"). Similarly, if 
you know that your 25-year old female neighbour from Utrecht participated in 
this study, you learn that she suffers from cancer ("background knowledge attack").

```{r l-diversity-data, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)

# Sample data frames
df3 <- data.frame(Nr = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), 
                  Age = c("=< 20", "=< 20", "=< 20", "20-30", "20-30", "20-30",
                          "20-30",  "20-30",  "20-30",  "20-30"),
                  Sex = c("Male", "Male", "Male", "Female", "Male", "Male", 
                          "Male", "Female", "Female", "Female"),
                  City = c("Rotterdam", "Rotterdam", "Rotterdam", "Rotterdam", 
                           "Zwolle", "Zwolle", "Zwolle", "Utrecht", 
                           "Rotterdam",  "Utrecht"),
                  Disease = c("Viral infection", "Heart-related", "Cancer", 
                              "Viral infection", "No illness", "Tuberculosis", 
                              "Heart-related", "Cancer", "Viral infection", 
                              "Cancer"))

df4 <- data.frame(Nr = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10), 
                  Age = c("=< 20", "=< 20", "=< 20", "20-30", "20-30", "20-30",
                          "20-30",  "20-30",  "20-30",  "20-30"),
                  Sex = c("Male", "Male", "Male", "Female", "Male", "Male", 
                          "Male", "Female", "Female", "Female"),
                  City = c("Rotterdam", "Rotterdam", "Rotterdam", "*", 
                           "Zwolle", "Zwolle", "Zwolle", "*", 
                           "*",  "*"),
                  Disease = c("Viral infection", "Heart-related", "Cancer", 
                              "Viral infection", "No illness", "Tuberculosis", 
                              "Heart-related", "Cancer", "Viral infection", 
                              "Cancer"))

# Convert data frames to tables
table3 <- 
  kbl(df3, caption = "2-anonymous dataset", label = NA) %>% 
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 11,
    #fixed_thead = list(enabled = T, background = "#FFCD00")
  ) %>%
  column_spec(2, bold = T) %>%
  row_spec(row = 1:3, background = "#ffedc3") %>%
  row_spec(row = c(4, 9), background = "#b2dfd8") %>%
  row_spec(row = 5:7, background = "#c9e0fc") %>%
  row_spec(row = c(8, 10), background = "#fad6bf") %>%
  footnote(general = "Colours indicate an 'equivalence class' of quasi-identifers", general_title = "", footnote_as_chunk = T)

table4 <- 
  kbl(df4, caption = "2-anonymous 2-diverse dataset", label = NA) %>% 
  kable_styling(
    full_width = FALSE,
    position = "left",
    font_size = 11 #,
    #fixed_thead = list(enabled = T, background = "#FFCD00")
  ) %>%
  column_spec(2, bold = T) %>%
  row_spec(row = 1:3, background = "#ffedc3") %>%
  row_spec(row = c(4, 8:10), background = "#b2dfd8") %>%
  row_spec(row = 5:7, background = "#c9e0fc") %>%
  footnote(general = "Colours indicate an 'equivalence class' of quasi-identifers and sensitive attributes", 
           general_title = "", footnote_as_chunk = T)


if(knitr::is_html_output()){
cat(c(
      '<table><tr valign="top"><td>', 
      table3, 
      '</td><td>', 
      table4, 
      '</td><tr></table>'),
    sep = '')
  
} else if(knitr::is_latex_output()){
    print(table3)
    print(table4)
} else if(knitr::is_epub_output()){
  print(table1)
  print(table2)
}
```

K-anonymity does not protect against such 
[homogeneity and background knowledge attacks](https://doi.org/10.1145/1217299.1217302){target="_blank"}. 
Therefore, L-diversity proposes that there should be at least L different values 
for the sensitive attribute per combination of quasi-identifiers. In the example 
above, if we choose an L of 2, that means that for each combination of Age, Sex 
and City, there are at least 2 distinct diseases. In the example, we suppressed 
City for these homogeneous cases, so that all females between 20 and 30 years 
old can either have cancer or a viral infection.

Like k-anonymity, there is 
[no perfect value of L](https://desfontain.es/privacy/l-diversity.html#the-bad-news-policy){target="_blank"}, 
although it is usually less or equal to k and more than 1.

The below [video](https://www.youtube.com/watch?v=GNhb3PcmjmA){target="_blank"} 
explains the concept of L-diversity using an example: 

<iframe width="100%" height="400" src="https://www.youtube-nocookie.com/embed/GNhb3PcmjmA?start=24&end=151" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>

#### T-closeness {#t-closeness}

T-closeness ensures that the distribution of a *sensitive attribute* within a 
generalisation of a *quasi-identifier* is close to the distribution of the 
sensitive attribute in the entire dataset. In other words, it ensures that the 
sensitive attribute is not skewed towards a specific value within a group of 
similar individuals, which could potentially be used to re-identify someone. 
For example, if a dataset contains information on Age (quasi-identifier), Sex 
(quasi-identifier), and Income (sensitive attribute), and t-closeness is applied 
with a value of t = 0.1, then for each combination of Age and Sex, the 
distribution of income must be within 10% of the distribution of income in the 
entire dataset. 

T-closeness can get complicated quite fast. If you’re curious to know how it 
works, the below [video](https://www.youtube.com/watch?v=Upb8jqlsbFM){target="_blank"} 
explains the concept of t-closeness using an example: 

<iframe width="100%" height="400" src="https://www.youtube-nocookie.com/embed/Upb8jqlsbFM?start=23&end=404" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>

### When to use {#when-k-anonymity}

K-anonymity, L-diversity and t-closeness are usually applied to de-identify 
tabular datasets, before being shared. They are also most suitable for 
relatively large datasets (i.e., containing a large number of individuals), as 
more details (utility) are likely to be retained in such datasets
([source](https://www.duo.uio.no/bitstream/handle/10852/79902/13/Anonymization-of-Health-Data.pdf#page=130){target="_blank"}). 

### Implications for research {#implications-k-anonymity}

- It is very easy to lose a lot of the (granularity of the) data when satisfying 
the k-, L- or T-criteria: the higher the criteria, the lower the risk of 
re-identification, but the more information you lose. The balance between 
privacy and utility is therefore very important to take into consideration when 
applying these privacy models. 
- The more variables (quasi-identifiers), the larger the dataset and the more 
outliers there are in the dataset, the more difficult de-identification will be 
without losing too much information 
([as shown here](https://programming-dp.com/ch2.html){target="_blank"}).  
- If a dataset is k-anonymous, L-diverse or T-close, that does not mean that 
the dataset is also considered [anonymous under the GDPR](#when-anonymous). The 
degree of anonymity after applying these approaches depends entirely on your 
own choices in terms of k, L or T, in terms of the variables that you included, 
and on the context of your dataset. For example, if you failed to include a 
quasi-identifier in k-anonymising your dataset, your dataset is in reality not 
k-anonymous. 

### Further reading {#resources-k-anonymity}

- Read the original papers about k-anonymity 
([Sweeney, 2002](https://doi.org/10.1142/S0218488502001648){target="_blank"}, 
[pdf](https://dataprivacylab.org/dataprivacy/projects/kanonymity/kanonymity.pdf){target="_blank"}), 
L-diversity ([Machanavajjhala et al., 2007](https://doi.org/10.1145/1217299.1217302){target="_blank"}) 
and t-closeness ([Li et al., 2007](https://doi.org/10.1109/ICDE.2007.367856){target="_blank"}, 
[pdf](https://www.cs.purdue.edu/homes/ninghui/papers/t_closeness_icde07.pdf){target="_blank"}). 
- In response to k-anonymity, L-diversity and T-closeness, other approaches have 
been formulated as well, such as k-Map, delta-presence (δ), beta-likeness (β) and 
delta-disclosure (δ). We do not go into those here, but 
[Damien Desfontaines’s blogposts](https://desfontain.es/privacy/archives.html){target="_blank"} 
and [this thesis](http://hdl.handle.net/10852/79902){target="_blank"} 
([pdf](https://www.duo.uio.no/bitstream/handle/10852/79902/13/Anonymization-of-Health-Data.pdf#page=44){target="_blank"}) 
discuss some of these approaches in more detail. 
- On [this page](https://elf11.github.io/2017/04/22/kanonymity.html){target="_blank"} 
and [here](https://programming-dp.com/ch1.html){target="_blank"} you can find 
more information on the different attacks that can potentially be committed to 
datasets. 

## Differential privacy {#differential-privacy}

:::keywords
On this page: differential privacy, dp, epsilon, quantifying
privacy, privacy-utility spectrum
Date of last review: 2023-05-30
:::

Differential privacy (DP) is a mathematical technique to prevent the disclosure 
of personal information by adding statistical noise to queries. In its original 
form, the data are stored on a secure server. Researchers can then, without 
access to the data, pose a query (i.e., an analysis request), and obtain the 
result with a certain amount of statistical noise added to the output. The 
amount of noise is tweaked such that adding or removing an individual to or from 
the dataset alters the result of the query by a limited, predetermined amount.  

The below [video](https://www.youtube.com/watch?v=gI0wk1CXlsQ){target="_blank"} 
explains the basic premise of differential privacy using a very basic example:  

<iframe width="100%" height="400" src="https://www.youtube-nocookie.com/embed/gI0wk1CXlsQ?start=185" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>

:::note
Differential privacy is generally not added to a dataset itself, but to the 
**result of a query** (e.g., average income in a dataset), although some forms 
exist in which the dataset itself is altered; this is called *local* 
differential privacy. 
:::

### How it works

Imagine a dataset with peoples' incomes, and we want to know the average income 
in the data (i.e. our query). When applying differential privacy, we want to 
ensure that we cannot determine whether an individual is in the dataset or not. 
In theory, we can do this by adding or removing any single individual to or from 
our dataset and recalculating the average income in the modified dataset. 
Importantly, additions to the dataset should all be individuals that *could be* 
in the dataset, and not only those that already are. Depending on the privacy 
requirements (i.e., what amount of privacy leakage is acceptable) and the 
sensitivity of the query (the effect of changing one individual on the average 
income), more or less statistical noise is added to the result. 

The most common definition of differential privacy is "epsilon-differential 
privacy", epsilon being the "**privacy budget**". In this definition, we 
consider two scenarios: the original dataset and a modified dataset, where we 
substitute the person with the highest income with the person with the lowest 
income. A certain amount of noise is added, so that the probabilities of 
obtaining a certain result in the original and the modified datasets do not 
differ more than Epsilon. Epsilon (i.e., the privacy budget) is a predetermined 
value: the lower epsilon, the higher the level of privacy, and therefore the 
more noise is added to the query. 

:::fyi
Differential privacy can also protect *groups* of people by adding or 
subtracting groups instead of individuals. This can for example be useful 
when you want to protect a household as a whole, and not only the individuals 
in it. 
:::

:::warning
Setting a lower epsilon results in more noise in the results that you release. 
This simply implies that it becomes more difficult to distinguish whether 
individuals were or were not in your dataset. However, it does **not** necessarily 
imply that a certain level of epsilon will result in fully anonymous data. 
Whether a dataset can be viewed as anonymous or not under the GDPR will always 
be context- and dataset-dependent. 
:::

### Implications for research {#dp-implications}

- The advantage of differential privacy is that privacy risk can be 
mathematically quantified (through epsilon). With many other techniques, it can 
be hard to determine exactly what the privacy guarantees are, especially with 
the presence of external information. 
- As with the other statistical approaches, there is a trade-off between more 
privacy or more utility: the lower the privacy budget, the more privacy is 
retained, but also the less accurate (more noisy) individual values will be. 
- The privacy budget should be determined in advance and can only be spent once: 
every query will lower the available privacy budget by epsilon, and when the 
privacy budget reaches zero, no more queries can be done. Thus, **you cannot 
indefinitely reuse the data and still preserve privacy**. 
- A disadvantage of differential privacy is that the concept is not completely 
trivial for the non-expert and in some cases, this has resulted in 
[violated privacy guarantees](https://doi.org/10.48550/arXiv.2011.07018){target="_blank"}. 
There is **no absolute right way to set the privacy budget** and no framework 
to decide which value of epsilon should be set for what kind of data. Thus, it 
can be hard for researchers to justify using a particular value of epsilon.

### When to use {#dp-when}

The implementation of differential privacy is a very technical (and tricky) 
endeavour. Setting up a differentially private server that only outputs 
differentially private results for any query is currently practically impossible. 
Thus, if you are not an expert on statistics and/or differential privacy, we 
recommend reaching out to someone who is, and to use Differential Privacy in 
addition to other privacy-enhancing techniques, such as pseudonymisation.

For now, differential Privacy is mostly used in combination with synthetic data, 
or with simple repetitive queries. Widespread use of differential privacy might 
become safer and/or easier over time as implementations are tested more 
thoroughly on real-world datasets. 

### Further reading {#dp-resources}

- Damien Desfontaines has written a number of easy to read 
[blogposts about differential privacy](https://desfontain.es/privacy/friendly-intro-to-differential-privacy.html){target="_blank"}, 
including [this one](https://desfontain.es/privacy/differential-privacy-awesomeness.html){target="_blank"} 
explaining the basic premise of differential privacy. 
- The online book [Programming Differential Privacy](https://programming-dp.com/index.html){target="_blank"} 
explains in detail how differential privacy works. 
- Or read [Differential privacy: a primer for the non-technical audience](https://privacytools.seas.harvard.edu/files/privacytools/files/pedagogical-document-dp_new.pdf){target="_blank"}.
- Background information: 
[The Algorithmic Foundations of Differential Privacy](https://www.cis.upenn.edu/~aaroth/Papers/privacybook.pdf){target="_blank"}.
